### 2017-07-06 inference.tex
Sec. 5 weakening example.

Fig.~\ref{fig:inference-rules} shows the syntax-directed inference
rules for $\txnimp$ commands inside a transaction $\C{TXN}_i$.
Inference depends on the $\I$-constrained rely relation $\R$, and the
high-level invariant $I$ for the reasons that will become clear
shortly.  Inference rules compute, for each command $c$, a (meta)
function $\F$ that returns a set as an expression in $\SL$, given a
pair of sets ($\stl$ and $\stg$) that describe local and a global
database states, respectively. The expression returned by
$\F(\stl,\stg)$ (abstractly) describes the set of records that get
added to $\stl$ as a result of executing $c$ under $\stl$ and $\stg$.
Thus, $\F$ captures the \emph{effect} part of the state transformer of
$c$, which is the function $\lambda(\stl,\stg).~\stl \cup
\F(\stl,\stg)$. For $\F$ to be useful in RG verification, it needs to
be stable w.r.t the rely relation $\R$. The stability condition on
effects is defined below:
\begin{smathpar}
\begin{array}{lcl}
  \stable(\R,\F) & \Leftrightarrow & \forall \stl,\stg,\stg'.~
  \R(\stl,\stg,\stg') \Rightarrow \F(\stl,\stg) \equiv \F(\stl,\stg')
\end{array}
\end{smathpar}
Intuitively, the stability condition requires an effect to describe
the same set of records before and after the interference. The
equivalence in $\SL$ translates to the equality in first-order logic,
as will be explained shortly. In the inference rules, stability is
enforced constructively by a meta function $\stabilize{\cdot}$, which
accepts an effect and returns a new effect that is guaranteed to be
stable under $\R$. $\stabilize{\cdot}$ achieves the stability
guarantee by abstracting away the bound $\stg$ in an unstable $\F$ to
an existentially bound $\stg'$ as described below:
\begin{smathpar}
\begin{array}{lcll}
  \stabilize{\F} & = & \F & \texttt{if } \stable(\R,\F).\\
  & = & \lambda (\stl,\stg).~\existsl(\stg',I(\stg'),\F(\stl,\stg')) 
      & \texttt{otherwise. }\stg'\texttt{ is a fresh name.}\\
\end{array}
\end{smathpar}
Observe that when $\F$ is not stable, then $\stabilize{\F}$ returns an
$\F'$ that simply ignores its $\stg$ argument in favor of a generic
$\stg'$, making $\F'$ trivially stable. The existentially bound
$\stg'$ satisfies $I$ since transactions are always executed against a
valid database state.

>>>>

Fig.~\ref{fig:inference-rules} shows the syntax-directed inference
rules for $\txnimp$ commands inside a transaction $\C{TXN}_i$.
Inference depends on the $\I$-constrained rely relation $\R$, and the
high-level invariant $I$ for the reasons that will become clear
shortly.  Inference rules compute, for each command $c$, a (meta)
function $\F$ that returns a set as an expression in $\SL$, given a
pair of sets ($\stl$ and $\stg$) that describe local and a global
database states, respectively. The expression returned by
$\F(\stl,\stg)$ (abstractly) describes the set of records that get
added to $\stl$ as a result of executing $c$ under $\stl$ and $\stg$.
Thus, $\F$ captures the \emph{effect} part of the state transformer of
$c$, which is the function $\lambda(\stl,\stg).~\stl \cup
\F(\stl,\stg)$. For $\F$ to be useful in RG verification, it needs to
be stable w.r.t the rely relation $\R$. The stability condition on
effects is defined below:
\begin{smathpar}
\begin{array}{lcl}
  \stable(\R,\F) & \Leftrightarrow & \forall \stl,\stg,\stg',\bar{v}.~
  \R(\stl,\stg,\stg') \Rightarrow \F(\stl,\stg) \equiv \F(\stl,\stg')
\end{array}
\end{smathpar}
where, $\bar{v}$ are the variables that occur free in $\F$ (possible
because of how the inference rules are structured).  Intuitively, the
stability condition requires an effect to describe the same set of
records before and after the interference. The equivalence in $\SL$
translates to the equality in first-order logic, as will become clear
shortly. In the inference rules, stability is enforced constructively
by a meta function $\stabilize{\cdot}$, which accepts an effect and
returns a new effect that is guaranteed to be stable under $\R$.
$\stabilize{\cdot}$ achieves the stability guarantee by abstracting
away the bound $\stg$ in an unstable $\F$ to an existentially bound
$\stg'$ as described below:
\begin{smathpar}
\begin{array}{lcll}
  \stabilize{\F} & = & \F & \texttt{if } \stable(\R,\F).\\
  & = & \lambda (\stl,\stg).~\existsl(\stg',I(\stg'),\F(\stl,\stg')) 
      & \texttt{otherwise. }\stg'\texttt{ is a fresh name.}\\
\end{array}
\end{smathpar}
Observe that when $\F$ is not stable, then $\stabilize{\F}$ returns an
$\F'$ that simply ignores its $\stg$ argument in favor of a generic
$\stg'$, making $\F'$ trivially stable. It is safe to assume
$I(\stg')$ because all verified transactions preserve the invariant,
hence only valid database states will ever be witnessed. From the
perspective of RG reasoning, $\stabilize{\cdot}$ effectively weakens
the post-condition of a statement, as done by the
\rulelabel{RG-Conseq} rule for transaction-bound commands (cf.  the
supplementary material). The weakening semantics chosen by
$\stabilize{\cdot}$, while being simple, is nonetheless useful because
of the $I(\stg')$ assumption on the existentially bound $\stg'$. The
example in Fig.~\ref{fig:weakening-example} demonstrates. 
\begin{figure}[h]
\begin{ocaml}
let add_interest acc_id pc = atomically @@ do
  let a = SQL.select1 BankAccount (fun acc -> acc.id = acc_id);
  let y = a.bal + pc*a.bal;
  SQL.update BankAccount (fun acc -> {acc with bal = acc.bal + y})
                         (fun acc -> acc.id = acc_id)
\end{ocaml}
\caption{A transaction that deposits an interest to a bank account}
\label{fig:weakening-example}
\end{figure}
Here, an \C{add\_interest} transaction adds a positive interest to the
balance of a bank account, which is required to be positive ($I(\stg)
\Leftrightarrow \forall(r\in\stg).~r.\C{bal}\ge 0$). The transaction
starts by issuing a \C{select1} query, whose effect implicitly asserts
that there exists a record in the bank account database that is equal
to $a$ ($\exists(r\in\stg).~r=a$). However, this assertion is unstable
because a concurrent \C{withdraw} or \C{deposit} transactions might
update the account balance, and there will no longer exist an
$r\in\stg$ that is equal to $a$. Fortunately, the weakening operator
($\stabilize{\cdot}$) comes to the rescue here by weakening the
assertion to $\exists \stg'.I(\stg') \conj \exists(r\in\stg').~a=r$,
which is enough to prove that $\C{a.bal + pc*a.bal}\ge 0$, and verify
that \C{add\_interest} preserves the positive balance invariant.
